---
title: Theological Classicism in Revolutionary America
description: "Final project for Clio 2"
author:
  - name: Kris Stinson 
    url: https://visualizinghistory.org/
    affiliation: GMU CHNM
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(quanteda)
library(readtext)
library(stm)
library(Matrix)
library(broom)
library(stringr)
library(LDAvis)
library(dplyr)
library(tidytext)
library(ggplot2)
library(tm)
```

Ancient history was very important to Americans of the late eighteenth and early nineteenth century. Seeped in a pervasive historical culture largely inherited from England, Americans interpreted modern events through the lens of antiquity before, during, and after the American Revolution. Attempting to capture American's fascination with ancient history during this era, historians have described this historical culture as "the culture of classicism" and the "cult of antiquity." (Richard, Winterer, Reinhold) This fixation with classical antiquity defined the intellectual atmosphere of early American colleges, making higher education during this time, if not learning itself, inherently classical. (Winterer) As such, much work has been done analyzing the many political pamphlets, speeches, and other published texts that have made explicit connections and reference to hallowed antiquity. 

However, as other scholars have shown, the individuals who comprised England’s North American colonies were inheritors of multiple, often conflicting intellectual traditions. Specifically, these individuals were descendants of religious traditions, mainly Protestant, that were equally useful and applicable to contemporary events. (Pestana, Perl-Rosenthal, Shalev) Additionally, the spokespeople of this tradition, mainly the ministers and preachers of local congregations, had themselves come from many of the early colleges and, as a result, had been cast in the classical mold. Indeed, it is perhaps useful to keep in mind that, for many early colleges in America, training future clergy was the primary function of the school.  And, like any committed student, they did not forget what they had learned once they had left the classroom behind. The classical concepts, characters, and rhetoric went with them and became woven into the content of their sermons; making the average sermon perhaps the most effective way in which the culture of classicism was spread throughout early America to people and places beyond the reach of the colleges. In analyzing the fifty-five sermons within Liberty Fund’s *Political Sermons of the Founding Era: 1730-1805*, one is able to begin to glimpse the presence of this “theological classicism,” for in their attempt to comprehend the nature of their God and religion’s relation to human governments, ministers turned to classical antiquity. 

Yet any study that would attempt to mine for classicism in a set of late eighteenth sermons seems destined for confusion. For ministers of the Revolutionary era, "Rome" was an especially loaded term. Comprised in an atmosphere of heightened anti-Catholicism, came to refer to the seat of the "Man of Sin" and all the evils of popery. One might expect, then, many of the allusions to Rome within these sermons would largely be in reference to the dangers of Catholic influence and not the ancient civilization. However, a cursory review of the sermons revealed that this is not the case. While Rome’s connection to Catholicism is definitely present, equally so are detailed references to antiquity that often go far beyond mere lip service. 


```{r, fig.cap="This figure displays the prevalence of Rome as a concept throughout all fifty-five sermons. As was the case with 'Rome,' 'Roman' and 'Romans' were equally loaded terms since it was possible for the former to be in reference to Roman Catholics and the latter the biblical book of the same name. However, in both cases, references to antiquity outnumbered those to more religious topics. Only twelve of forty-six uses of 'Roman' were in reference to Catholicism, while, more surprisingly, discussions on 'Romans' within these sermons were almost exclusively on the ancient society itself and not on the letter from St. Paul."}
my_csv <- read_csv("sermons.csv")
my_corpus <- corpus(my_csv)

classics <- c("Rome", "Roman", "Romans", "Greece", "Greek", "Greeks","Sparta", "Athens", "Virgil", "Cicero", "Livy",
              "Plutarch", "Cato", "Socrates", "Caesar", "Cesar", "Thucydides", 
              "Demosthenes", "Sallust", "Homer", "Tacitus", "Dionysius", 
              "Polybius", "Ovid", "Seneca", "Horace", "Plato", "Sylla", "Marius", "Aristotle")


Rome <- c("Rome", "Roman", "Romans")
Greece <- c("Greece", "Greek", "Greeks", "Sparta", "Athens")
classical_characters <- c("Sylla", "Marius", "Virgil", "Cicero", "Livy",
                       "Plutarch", "Cato", "Socrates", "Caesar", "Cesar", "Thucydides", 
                       "Demosthenes", "Sallust", "Homer", "Tacitus", "Dionysius", 
                       "Polybius", "Ovid", "Seneca", "Horace", "Plato", "Aristotle")

kwic(my_corpus, pattern = Rome) %>%
  textplot_xray() +
  aes(color = keyword)+ 
  scale_color_manual(values = c("purple", "gold", "black")) +
  theme(legend.position = "none") +
  labs(title= "Rome in the Sermons", y = "Sermon")

```
```{r, fig.cap="Representing the other half of the American classical mind, this figure shows the prevalence of ancient Greece throughout the fifty-five sermons. It illustrates that although the topic of Greece itself was not as popular as Rome, that did not make it any less present. Not only do many of these occurrences coincide with discussions of Rome but rival the amount of references to some sacred places, such as Jerusalem."}
kwic(my_corpus, pattern = Greece) %>%
  textplot_xray() +
  aes(color = keyword)+ 
  scale_color_manual(values = c("blue", "black", "gold", "red", "green")) +
  theme(legend.position = "none")+
  labs(title = "Ancient Greece in the Sermons", y = "Sermon")
```

The classicism exhibited by these political sermons went far beyond the more general references to the ancient civilizations as a whole. Frequently, specific classical characters themselves were called out and employed as a means to illustrate the character of God and the truth and effectiveness of the gospel to political society. To get a sense as to how comprehensive this trend was, I have cross-referenced the *Political Sermons* with Meyer Reinhold’s *The Classick Pages*, a study into some of the most popular classical works in Revolutionary America. Using Reinhold’s study, I constructed a dictionary of the most widely read works that comprised the classical understanding of the late eighteenth and early nineteenth centuries. After the initial comparison, one is surprised with how completely this small sample of early American sermons covered the gamut of classical references. Indeed, in the list of popular classical authors and characters included in Reinhold’s study, the *Political Sermons of the American Founding Era* contained nearly all of them. ^[The only names the sermons omitted were Ovid, Polybius, and, more surprisingly, Livy.]

```{r}
mystopwords <- tibble(word = c("g", "i'll", "em", "d", "ye", "mr", "i've", "tis", "O", "o"))
sermon_words <- my_csv %>% 
  unnest_tokens(word, text) %>%
  count(meta, word, sort = TRUE)

total_words <- sermon_words %>% 
  group_by(meta) %>% 
  summarize(total = sum(n))

sermon_words <- left_join(sermon_words, total_words)
rm(total_words)

freq_by_rank <- sermon_words %>% 
  group_by(meta) %>% 
  mutate(rank = row_number(), 
         `term frequency` = n/total)

sermon_words <- sermon_words %>%
  bind_tf_idf(word, meta, n)

sermon_words <- sermon_words %>%
  anti_join(mystopwords) %>% 
  select(-total) %>%
  arrange(desc(tf_idf))

classics_words <- sermon_words %>% 
  filter(str_to_lower(word) %in% str_to_lower(classics)) %>% 
  arrange(desc(n))

classics_words %>% count(word) %>% 
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  labs(title = "Classical References", x = "Word", y = "Number of Sermons") +
  geom_col() + 
  coord_flip()
```

Ministers of the Revolutionary period ultimately displayed a deep sense of classicism. Caroline Winterer has made a distinction between those who possessed “thick” and “thin” classicism: those who had “little formal learning” of antiquity and possessed a thin understanding of the classics, and those who possessed a thick knowledge from a “deep, lifelong study” of classicism. ^[Caroline Winterer, "Model Empire, Lost City: Ancient Carthage and the Science of Politics in Revolutionary America," *The William and Mary Quarterly*, Vol. 67. No. 1 (Omohundro Institute of Early American History and Culture, 2010), 23-24.]

```{r}
doc42_44 <- my_csv %>% 
  filter(doc_id %in% c("doc42", "doc44"))
cs <- readtext("common_sense.txt")
aor <- readtext("age_of_reason.txt")

cs_corpus <- corpus(cs)
aor_corpus <- corpus(aor)

cs_words <- cs %>% 
  unnest_tokens(word, text) %>%
  count(doc_id, word, sort = TRUE)
aor_words <- aor %>% 
  unnest_tokens(word, text) %>%
  count(doc_id, word, sort = TRUE)


total_words <- cs_words %>% 
  group_by(doc_id) %>% 
  summarize(total = sum(n))

cs_words <- left_join(cs_words, total_words)
rm(total_words)

total_words <- aor_words %>% 
  group_by(doc_id) %>% 
  summarize(total = sum(n))

aor_words <- left_join(aor_words, total_words)
rm(total_words)

freq_by_rank <- cs_words %>% 
  group_by(doc_id) %>% 
  mutate(rank = row_number(), 
         `term frequency` = n/total)

cs_words <- cs_words %>%
  bind_tf_idf(word, doc_id, n)
aor_words <- aor_words %>%
  bind_tf_idf(word, doc_id, n)

cs_words <- cs_words %>% 
  select(-total) %>%
  arrange(desc(tf_idf))
aor_words <- aor_words %>% 
  select(-total) %>%
  arrange(desc(tf_idf))

cs_classics_words <- cs_words %>% 
  filter(str_to_lower(word) %in% str_to_lower(classics)) %>% 
  arrange(desc(n))
aor_classics_words <- aor_words %>% 
  filter(str_to_lower(word) %in% str_to_lower(classics)) %>% 
  arrange(desc(n))

comp <- bind_rows(doc42_44, cs, aor)

comp_words <- comp %>%
  unnest_tokens(word, text) %>%
  count(doc_id, word, sort = TRUE)

comp_total_words <- comp_words %>% 
  group_by(doc_id) %>% 
  summarize(total = sum(n))

comp_words <- left_join(comp_words, comp_total_words)
rm(comp_total_words, book_words)


comp_classics_words <- comp_words %>% 
  filter(str_to_lower(word) %in% str_to_lower("Romans")) %>% 
  arrange(desc(n))

ggplot(comp_classics_words, aes(x = fct_reorder(word, n), y = n, fill=doc_id)) +
  labs(title="The Classicism of Ministers and Thomas Paine", x="Word", y="Number of References")+
  scale_fill_manual(labels = c("Paine's Age of Reason", "Paine's Common Sense", "Edwards Jr.'s Sermon", "Webster's Sermon"), values = c("blue", "black", "purple", "gold"))+
  geom_bar(stat="identity", position=position_dodge())
```
